Sparse spectral and $p$-finite element methods for partial differential equations on disk-slices and trapeziums

### 2019-05-13 papers/halfdisk/halfdisk.tex
Further edits

For disk slices and trapeziums, we can specialise the framework from Proposition \ref{prop:construction} in the following definition.
\begin{definition}\label{def:construction}
Let
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
with $\rho : (\alpha, \beta) \to \R$. Further, let $\genjacw^{(a,b,c)}(x)$ and $w^{(a,b)}_P(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align*}
\begin{cases}
\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
w^{(a,b)}_P(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align*}
where the three-parameter family of orthogonal polynomials on $[\alpha,\beta]$ given by $\{\genjac_n^{(a,b,c)}\}$ are \sotodo{Make clear whether they are orthonormal or not} orthogonal with respect to $\genjacw^{(a,b,c)}(x)$, and the two-parameter family of orthogonal polynomials on $[\gamma,\delta]$ given by $\{P_n^{(a,b)}\}$ are orthogonal with respect to $w^{(a,b)}_P(x)$.

Assume that either of the following hold:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(w^{(a,b)}_P\) is an even function (i.e. $a = b$, and we can hence denote the weight as $w^{(a)}_P(x) = (\delta-x^2)^a$).
\end{enumerate}
We define the four-parameter 2D orthogonal polynomials on $\Omega$ by
\begin{align*}
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: P_k^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align*}
which are orthogonal with respect to the weight
\begin{align*}
	W^{(a,b,c,d)}(x,y) := \genjacw^{(a,b,c+d)}(x) \: w_P^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega. 
\end{align*}
\end{definition}

@@

The two 1D families of orthogonal polynomials we have denoted $\{\genjac_{n}^{(a,b)}\}$ and $\{P_n^{(b,b)}\}$ are not orthonormal, but have (squared) norms given by the integral of their respective weight functions we denote as
\begin{align*}
	\normgenjac^{(a,b)} := \int_0^1 \: \genjacw^{(a,b)}(x) \: \D x, \quad \normjac^{(b)} := \int_{-1}^1 \: w_P^{(b)}(y) \: \D y,
\end{align*}
i.e. $\norm{\genjac_{n}^{(a,b)}}_{\genjacw^{(a,b)}}^2 = \normgenjac^{(a,b)}$ $\forall n = 0,1,2,\dots$ and $\norm{P_{n}^{(b,b)}}_{w_P^{(b)}}^2 = \normjac^{(b)}$ $\forall n = 0,1,2,\dots$. For clarity, we use the notation that 
\begin{align*}
	&\ip<p, q>_{\genjacw^{(a,b)}} := \int_0^1 p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x, \quad
		\ip<p, q>_{w_P^{(b)}} := \int_{-1}^1 p(y) \: q(y) \: w_P^{(b)}(y)\: \D y, \\
	&\norm{p}_{\genjacw^{(a,b)}}^2 := \ip<p, p>_{\genjacw^{(a,b)}}, \quad
		\norm{p}_{w_P^{(b)}}^2 := \ip<p, p>_{w_P^{(b)}}.
\end{align*}
Further, we have that the norms of the 2D half-disk OPs for a particular parameter family $(a,b)$ are then given by
\begin{align}
	\norm{\hdopnkab}_\Wab^2 &=  \iint_\Omega \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: \rho(x)^{2k} \: {P_k^{(b,b)}\fpr(\frac{y}{\rho(x)})^2} \: \Wab(x,y) \: \D y \: \D x \nonumber \\
	&= \iint_\Omega \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: {P_k^{(b,b)}\fpr(\frac{y}{\rho(x)})^2} \: \genjacw^{(a,b+k)}(x) \: w_P^{(b)}\fpr(\frac{y}{\rho(x)}) \: \D y \: \D x \nonumber \\
	&= \Big( \int_\alpha^\beta \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: \genjacw^{(a,b+k+\half)}(x) \: \D x \Big) \: \Big( \int_\gamma^\delta \: {P^{(b,b)}}(t)^2 \: w_P^{(b)}(t) \: \D t \Big) \nonumber \\
	&= \normgenjac^{(a,b+k+\half)} \: \normjac^{(b)} \label{eqn:normhdop}
\end{align}
where we have used the change of variable $t = \frac{y}{\rho(x)}$. Again, for clairty we use the notation that $\ip<p, q>_{\Wab} := \iint_\Omega p(x,y) \: q(x,y) \: \Wab(x,y) \: \D y \: \D x$ and $\norm{p}_{\Wab}^2 := \ip<p, p>_{\Wab}$ .

@@

The incrementing and decrementing of parameters as seen here is analogous to other well known orthogonal polynomials families' derivatives, for example the Jacobi polynomials on the interval, as seen in the DLMF \cite[(18.9.3)]{DLMF}, and on the triangle \cite{olver2018recurrence}.

@@

\begin{proof}
We proceed with the case for the operator $D_y^{(a,b)}$ for partial differentiation by $y$. Since $\{\hdop^{(a, b+1)}_{m,j}\}$ for $m = 0,\dots,n-1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n-1$ polynomial, we can expand $\pddy \hdopnkab = \sum_{m=0}^{n-1} \sum_{j=0}^m c_{m,j} \: \hdop^{(a, b+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. We can use an integration-by-parts argument to show that the only non-zero coefficient of this expansion is when $m = n-1$, $j = k-1$. First, note that
\begin{align*}
	c_{m,j} = \frac{\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,j} >_{W^{(a, b+1)}}}{\norm{\hdop^{(a, b+1)}_{m,j}}^2_{W^{(a, b+1)}}}.
\end{align*}
Then, using the change of variable $t = \frac{y}{\rho(x)}$, we have that
\begin{align*}
	\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,j} >_{W^{(a, b+1)}} &= \ip< \genjacnmk^{(a,b+k+\half)}, \rho(x)^{k+j} \: \genjacmmj^{(a,b+1+j+\half)} >_{\genjacw^{(a,b+1)}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{j}^{(b+1,b+1)} >_{w_P^{(b+1)}} \\
	&= \ip< \genjacnmk^{(a,b+k+\half)}, \genjacmmj^{(a,b+1+j+\half)} >_{\genjacw^{(a,b+1+\half(k+j))}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{j}^{(b+1,b+1)} >_{w_P^{(b+1)}}.
\end{align*}
Now, integrating by parts, noting that the weight $w_P^{(b)}$ is a polynomial degree $2b$ and vanishes at the limits of the integral for positive parameter $b$, we have that
\begin{align*}
	\ip< P_k^{(b,b) \: \prime}, \: P_{j}^{(b+1,b+1)} >_{w_P^{(b+1)}} &= \ip< P_k^{(b,b) \: \prime}, \: w_P^{(b+1)} \: P_{j}^{(b+1,b+1)} >_1 \\
	&= \ip< P_k^{(b,b)}, \: \ddy [w_P^{(b+1)} \: P_{j}^{(b+1,b+1)}] >_1 \\
	&= \ip< P_k^{(b,b)}, \: P_{j}^{(b+1,b+1) \: \prime} >_{w_P^{(b+1)}} - 2 b \ip< P_k^{(b,b)}, \: y \: P_{j}^{(b+1,b+1)} >_{w_P^{(b)}} \\
	&= \ip< P_k^{(b,b)}, \: w_P^{(1)} \: P_{j}^{(b+1,b+1) \: \prime} -2b \:y \: P_{j}^{(b+1,b+1)} >_{w_P^{(b)}} 
\end{align*}
which is zero for $j < k-1$. Further,
\begin{align*}
	\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,k-1} >_{W^{(a, b+1)}} &= \ip< \genjacnmk^{(a,b+k+\half)}, \genjac_{m+1-k}^{(a,b+k+\half)} >_{\genjacw^{(a,b+k+\half)}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{k-1}^{(b+1,b+1)} >_{w_P^{(b+1)}},
\end{align*}
which is zero for $m < n-1$. Finally, we simply have that
\begin{align*}
	c_{n-1,k-1} &= \frac{1}{\normjac^{(b)}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{k-1}^{(b+1,b+1)} >_{w_P^{(b+1)}}.
\end{align*}

>>>>

For disk slices and trapeziums, we can specialise the framework from Proposition \ref{prop:construction} in the following definition.
\begin{definition}\label{def:construction}
Let
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
with $\rho : (\alpha, \beta) \to \R$. Further, let $\genjacw^{(a,b,c)}(x)$ and $w^{(a,b)}_P(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align*}
\begin{cases}
\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
w^{(a,b)}_P(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align*}
and define the associated inner products by: \bstodo{Is it best to define these ip s with the normalisation?}
\begin{align}
	\ip< p, \: q >_{\genjacw^{(a,b,c)}} &:= \frac{1}{\normgenjac^{(a,b,c)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjacw^{(a,b,c)}(x) \: \D x, \label{eqn:ipgenjac} \\
	\ip<p, q>_{w_P^{(a,b)}} &:= \frac{1}{\normjac^{(a,b)}} \:\int_\gamma^\delta p(y) \: q(y) \: w_P^{(a,b)}(y)\: \D y. \label{eqn:ipjac}
\end{align}
where
\begin{align*}
	\normgenjac^{(a,b,c)} := \int_\alpha^\beta \: \genjacw^{(a,b,c)}(x) \: \D x, \quad \normjac^{(a,b)} := \int_\gamma^\delta \: w_P^{(a,b)}(y) \: \D y.
\end{align*}
Further, denote the three-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b,c)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}), and the two-parameter family of orthonormal polynomials on $[\gamma,\delta]$ by $\{P_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipjac}).

Assume that either of the following hold:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(w^{(a,b)}_P\) is an even function (i.e. $a = b$, and we can hence denote the weight as $w^{(a)}_P(x) = (\delta-x^2)^a$).
\end{enumerate}
We define the four-parameter 2D orthogonal polynomials on $\Omega$ by
\begin{align*}
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: P_k^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align*}
which are orthogonal with respect to the weight
\begin{align*}
	W^{(a,b,c,d)}(x,y) := \genjacw^{(a,b,c+d)}(x) \: w_P^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega,
\end{align*}
that is,
\begin{align*}
	\ip< \hdopnk^{(a,b,c,d)}, \: \hdopmj^{(a,b,c,d)} >_{W^{(a,b,c,d)}} &= \normgenjac^{(a,b+c+d+2k+1)} \: \normjac^{(d,c)} \: \delta_{n,m} \: \delta_{k,j},
\end{align*}
where
$\ip< f, \: g >_W^{(a,b,c,d)} := \iint_\Omega \: f(x,y) \: g(x,y) \: W^{(a,b,c,d)}(x,y) \: \D y \: \D x$.
\end{definition}

@@

The two 1D families of orthogonal polynomials we have denoted $\{\genjac_{n}^{(a,b)}\}$ and $\{P_n^{(b,b)}\}$ are orthonormal, with respect to the inner products respectively given by:
\begin{align*}
	\ip< p, \: q >_{\genjacw^{(a,b)}} &:= \frac{1}{\normgenjac^{(a,b)}} \: \int_0^1 p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x, \\
	\ip<p, q>_{w_P^{(b)}} &:= \frac{1}{\normjac^{(b)}} \: \int_{-1}^1 p(y) \: q(y) \: w_P^{(b)}(y)\: \D y,
\end{align*}
where 
\begin{align*}
	\normgenjac^{(a,b)} := \int_0^1 \: \genjacw^{(a,b)}(x) \: \D x, \quad \normjac^{(b)} := \int_{-1}^1 \: w_P^{(b)}(y) \: \D y,
\end{align*}
i.e. $\norm{\genjac_{n}^{(a,b)}}_{\genjacw^{(a,b)}} = 1$ $\forall n = 0,1,2,\dots$ and $\norm{P_{n}^{(b,b)}}_{w_P^{(b)}} = 1$ $\forall n = 0,1,2,\dots$. For clarity, we use the notation that $\norm{p}_{\genjacw^{(a,b)}}^2 := \ip<p, p>_{\genjacw^{(a,b)}}$ and that $\norm{p}_{w_P^{(b)}}^2 := \ip<p, p>_{w_P^{(b)}}$.

Further, we have that the norms of the 2D half-disk OPs for a particular parameter family $(a,b)$ are then given by
\begin{align}
	\norm{\hdopnkab}_\Wab^2 &=  \iint_\Omega \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: \rho(x)^{2k} \: {P_k^{(b,b)}\fpr(\frac{y}{\rho(x)})^2} \: \Wab(x,y) \: \D y \: \D x \nonumber \\
	&= \iint_\Omega \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: {P_k^{(b,b)}\fpr(\frac{y}{\rho(x)})^2} \: \genjacw^{(a,b+k)}(x) \: w_P^{(b)}\fpr(\frac{y}{\rho(x)}) \: \D y \: \D x \nonumber \\
	&= \Big( \int_\alpha^\beta \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: \genjacw^{(a,b+k+\half)}(x) \: \D x \Big) \: \Big( \int_\gamma^\delta \: {P^{(b,b)}}(t)^2 \: w_P^{(b)}(t) \: \D t \Big) \nonumber \\
	&= \normgenjac^{(a,b+k+\half)} \: \normjac^{(b)} \label{eqn:normhdop}
\end{align}
where we have used the change of variable $t = \frac{y}{\rho(x)}$. Again, for clairty we use the notation that $\ip<p, q>_{\Wab} := \iint_\Omega p(x,y) \: q(x,y) \: \Wab(x,y) \: \D y \: \D x$ and $\norm{p}_{\Wab}^2 := \ip<p, p>_{\Wab}$ .

@@

The incrementing and decrementing of parameters as seen here is analogous to other well known orthogonal polynomial families' derivatives, for example the Jacobi polynomials on the interval, as seen in the DLMF \cite[(18.9.3)]{DLMF}, and on the triangle \cite{olver2018recurrence}.

@@

\begin{proof}
We proceed with the case for the operator $D_y^{(a,b)}$ for partial differentiation by $y$. Since $\{\hdop^{(a, b+1)}_{m,j}\}$ for $m = 0,\dots,n-1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n-1$ polynomial, we can expand $\pddy \hdopnkab = \sum_{m=0}^{n-1} \sum_{j=0}^m c_{m,j} \: \hdop^{(a, b+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. We can use an integration-by-parts argument to show that the only non-zero coefficient of this expansion is when $m = n-1$, $j = k-1$. First, note that
\begin{align*}
	c_{m,j} = \frac{\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,j} >_{W^{(a, b+1)}}}{\norm{\hdop^{(a, b+1)}_{m,j}}^2_{W^{(a, b+1)}}}.
\end{align*}
Then, using the change of variable $t = \frac{y}{\rho(x)}$, we have that
\begin{align*}
	\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,j} >_{W^{(a, b+1)}} &= \ip< \genjacnmk^{(a,b+k+\half)}, \rho(x)^{k+j} \: \genjacmmj^{(a,b+1+j+\half)} >_{\genjacw^{(a,b+1)}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{j}^{(b+1,b+1)} >_{w_P^{(b+1)}} \\
	&= \ip< \genjacnmk^{(a,b+k+\half)}, \genjacmmj^{(a,b+1+j+\half)} >_{\genjacw^{(a,b+1+\half(k+j))}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{j}^{(b+1,b+1)} >_{w_P^{(b+1)}}.
\end{align*}
Now, integrating by parts, noting that the weight $w_P^{(b)}$ is a polynomial degree $2b$ and vanishes at the limits of the integral for positive parameter $b$, we have that
\begin{align*}
	\ip< P_k^{(b,b) \: \prime}, \: P_{j}^{(b+1,b+1)} >_{w_P^{(b+1)}} &= \ip< P_k^{(b,b) \: \prime}, \: w_P^{(b+1)} \: P_{j}^{(b+1,b+1)} >_1 \\
	&= \ip< P_k^{(b,b)}, \: \ddy [w_P^{(b+1)} \: P_{j}^{(b+1,b+1)}] >_1 \\
	&= \ip< P_k^{(b,b)}, \: P_{j}^{(b+1,b+1) \: \prime} >_{w_P^{(b+1)}} - 2 b \ip< P_k^{(b,b)}, \: y \: P_{j}^{(b+1,b+1)} >_{w_P^{(b)}} \\
	&= \ip< P_k^{(b,b)}, \: w_P^{(1)} \: P_{j}^{(b+1,b+1) \: \prime} -2b \:y \: P_{j}^{(b+1,b+1)} >_{w_P^{(b)}} 
\end{align*}
which is zero for $j < k-1$. Further,
\begin{align*}
	\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,k-1} >_{W^{(a, b+1)}} &= \ip< \genjacnmk^{(a,b+k+\half)}, \genjac_{m+1-k}^{(a,b+k+\half)} >_{\genjacw^{(a,b+k+\half)}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{k-1}^{(b+1,b+1)} >_{w_P^{(b+1)}},
\end{align*}
which is zero for $m < n-1$. Finally, we simply have that \bstodo{Make sure correct ip definition is being used here}
\begin{align*}
	c_{n-1,k-1} &= \frac{1}{\normjac^{(b)}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{k-1}^{(b+1,b+1)} >_{w_P^{(b+1)}}.
\end{align*}
